Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2023-09-04 paper/markdown/manuscript.Rmd
Update of the manuscript
Update of the manuscript, figures and supplementary figures - using improved functionalities of the 'clustTools' package

The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares) and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical clustering and KMEANS algorithms. 
The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
The training subset observations were assigned to the clusters with an inverse distance-weighted 27-nearest neighbor classifier.

@@

To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to partition around medoids clustering in respect to a broad range of psychometry scores [@Drost2018; @Schubert2019] (__Supplementary Table S2__). 
Three mental clusters: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress), named after their key mental characteristic were identified in the training subset (__Supplementary Figure S3__). 
Subsequently, the mental cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
(training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`), 
comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).

>>>>

The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares), separation between clusters (average silhouette width) [@Rousseeuw1987] and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical clustering and KMEANS algorithms. 
The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
The training subset observations were assigned to the clusters with an inverse distance-weighted 27-nearest neighbor classifier.

@@

To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to partition around medoids clustering in respect to a broad range of psychometry scores [@Drost2018; @Schubert2019] (__Supplementary Table S2__). 
Three mental clusters: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress), named after their key mental characteristic were identified in the training subset (__Supplementary Figure S3__). 
Subsequently, the mental cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
(training: `r signif(semi_clust$stats$variance[1], 2)`, test: `r signif(semi_clust$stats$variance[2], 2)`), 
comparable average silhouette statistics 
(training: `r signif(semi_clust$stats$mean_sil[1], 2)`, test: `r signif(semi_clust$stats$mean_sil[2], 2)`) [@Rousseeuw1987], 
comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).

### 2023-09-04 paper/markdown/supplementary_material.Rmd
Update of the manuscript
Update of the manuscript, figures and supplementary figures - using improved functionalities of the 'clustTools' package

For semi-supervised clustering, diagnostic and performance of the clustering analysis, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized. 
UMAP (uniform manifold approximation and projection) was done with the _umap_ and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) packages [@Konopka2022; @McInnes2018]. 
Cross-distances between the training and test subsets of the study cohort and cross-distances between clusters in the training and test subsets of the study cohort were computed with _proxy_ [@Meyer2022].

@@

To find the optimal size of a training subset of the study cohort, random subsamples of the study dataset of varying sizes were investigated (`r paste(pow$sizes, collapse = ', ')` observations, 50 random draws per subsample size). 
For each random subsample, clustering tendency was assessed by Hopkins statistic (function `get_clust_tendency()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
For n = 250 observations, the median Hopkins statistic was 
`r pow$hopkins %>% filter(size == 250) %>% .$hopkins_stat %>% median %>% signif(2)` (interquartile range: 
`r pow$hopkins %>% filter(size == 250) %>% .$hopkins_stat %>% quantile(probs = c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' - ')`) indicative of good clustering tendency. 
The Hopkins statistic value for this subset was comparable with the value for the entire subset 
(`r signif(pca$clust_tendency$data$hopkins_stat, 2)`). 
For these reasons, 3/4 of the entire dataset or n = 
`r count(ptsd$dataset, partition)$n[1]` was considered as the adequate size of the training subset of the study cohort for a reproducible clustering analysis.

@@

Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance in the training (V = `r signif(semi_clust$variance$variance[1], 2)`) and test subset (V = `r signif(semi_clust$variance$variance[2], 2)`). 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `dist()`, package _proxy_; __Supplementary Figure S\@ref(fig:fig-cosine)B__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).

@@

_(A) Comparison of explanatory performance and reproducibility of various clustering algorithms in the training subset. The explanatory performance was measured as a fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares). The reproducibility was assessed by the rate of correct cluster assignment in 10-fold cross-validation (CV) with cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier. Note the superior stability of the PAM/cosine distance algorithm._

>>>>

For semi-supervised clustering, diagnostic and performance of the clustering analysis, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized. 
UMAP (uniform manifold approximation and projection) was done with the _umap_ and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) packages [@Konopka2022; @McInnes2018]. 
Cross-distances between the training and test subsets of the study cohort and cross-distances between clusters in the training and test subsets of the study cohort were computed with _philentropy_ [@Drost2018] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).

@@

To find the optimal size of a training subset of the study cohort, random subsamples of the study dataset of varying sizes were investigated (`r paste(pow$sizes, collapse = ', ')` observations, 50 random draws per subsample size). 
For each random subsample, clustering tendency was assessed by Hopkins statistic (function `get_clust_tendency()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Hopkins1954]. 
For n = 250 observations, the median Hopkins statistic was 
`r pow$hopkins %>% filter(size == 250) %>% .$hopkins_stat %>% median %>% signif(2)` (interquartile range: 
`r pow$hopkins %>% filter(size == 250) %>% .$hopkins_stat %>% quantile(probs = c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' - ')`) indicative of good clustering tendency. 
The Hopkins statistic value for this subset was comparable with the value for the entire subset 
(`r signif(pca$clust_tendency$data$hopkins_stat, 2)`). 
For these reasons, 3/4 of the entire dataset or n = 
`r count(ptsd$dataset, partition)$n[1]` was considered as the adequate size of the training subset of the study cohort for a reproducible clustering analysis.

@@

Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
training (variance = `r signif(semi_clust$stats$variance[1], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[1], 2)`) and 
test subset (variance = `r signif(semi_clust$stats$variance[2], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[2], 2)`). 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).

@@

_(A) Comparison of explanatory performance, cluster separation and reproducibility of various clustering algorithms in the training subset. The explanatory performance was measured as a fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares). The cluster separation was assessed with average silhouette width. The reproducibility was assessed by the rate of correct cluster assignment in 10-fold cross-validation (CV) with cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier. Note the superior stability of the PAM/cosine distance algorithm._
